Numerical modelling of the Cordilleran ice sheet

### 2014-07-14 kappa.tex
Re-think software implementation outline.
Begin to implement Arjen's comments.

% My commands
%\newcommand{\note}[1]{\textbf{[NOTE: #1]}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
\newcommand{\mref}[0]{\textbf{[ref.]}}

@@

% Document properties
\title{Numerical simulation of the Cordilleran ice sheet \\
       through the last glacial cycle}
%\runningtitle{Numerical modelling of the Cordilleran ice sheet}
\author{Julien Seguinot}

@@

\newpage
\maketitle
\tableofcontents
\newpage

@@

In our daily lives, we most commonly experience ice as the small, solid cubes
that crackle on contact with refreshing liquids. Yet, at geological
time-scales, and under high stresses, ice behaves in a much different fashion.
In the cold and mountainous regions of the globe, ice accumulates under the
action of cumulative snowfall, until it reaches a critical thickness, and
deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
Glaciers, these bodies of moving ice, flow by a combination of water-induced
sliding at the base \citep{Saussure.1796}, and visco-plastic deformation within
the ice body \citep{Forbes.1846b}.

As ice flows and glides on its bed, it transports rock debris and erodes the
landscape, thence leaving geomorphological traces of its former presence
(\cref{fig:glaciation-indicators}). In many parts of the world, montane people
and early explorers have since long learned to read these traces, and to
understand that glaciers had once been more extensive than today. With the
advent of the glacial theory, birthed the idea that, under colder temperatures
than today, expansive \emph{ice sheets} had once covered much of Europe and
North America \citep{Agassiz.1840}.

However, the glacial theory did not gain much belief until the exploration and
discovery of the two present-day ice sheets on Earth: the Greenland and
Antarctic ice sheets, provided a modern analogue for the proposed European and
North American ice sheets. The first traverse of Greenland was performed by
Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
the European Alps, it was then discovered from observations of the glacial
landscape, that there had not been only one, but at least four, glacial periods
\citep{Penck.Bruckner.1909}.

More recently, palaeo-climate records extracted from deep-sea sediments have
provided a much more detailed picture of the Earth environmental history
\citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973}, thus indicating neither
one, nor four, but rather tens of glacial cycles. During the last million
years, glacial and interglacial cycles have succeeded each other with a
100\,kyr periodicity \citep{Hays.etal.1976}. In the present thesis, we focus on
the most recent of these periods: \emph{the last glacial cycle}
(\cref{fig:paleo-timeseries})

% ice volume 26.07 + 2.93 = 29.00
% meters sea-level 58.3 + 0.70 = 59.0
Together, the Greenland and Antarctic ice sheets currently host most glacial
ice on the planet, corresponding to 59~meters of potential sea-level rise
\citep{Bamber.etal.2001, Fretwell.etal.2013}. At the apogee of the last glacial
cycle, the \emph{Last Glacial Maximum} (LGM), sea-level was 120 to 135\,m lower
than today
\citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
ice sheets and glaciers on land. Systematic mapping and compilation of glacial
landforms from all continents permitted reconstructions of this former ice
cover \citep{Ehlers.Gibbard.2007}, indicating that ice volume
was repartitioned between more extensive Greenland and Antarctic ice sheets,
the Scandinavian and Barents-Kara ice sheets in the Eurasian arctic, and the
Innuitian and Laurentide ice sheets and, finally, the \emph{Cordilleran ice
sheet}, over North America (\cref{fig:paleo-glaciation}).

Nevertheless, the landscape record is sparse in time and most often spatially
incomplete. Palaeo-ice sheets did not leave a systematic record of their
passage, and much of the evidence has been overridden by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can help.
Although these models alone contain too many uncertainties to reproduce
accurate picture of past glacial cycles \citep[e.g.,][]{Hebeler.etal.2008},
they can be calibrated against geological data where they exist, while
allowing a physics-based, quantitative reconstruction where they do not
\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.

@@

% ======================================================================
\section{Field area description}
% ======================================================================

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, \emph{the Cordilleran ice sheet} capped the mountain
ranges of the Western Cordillera, which forms our region of focus in the
present thesis.

@@

% ----------------------------------------------------------------------
\subsection{Topographic setting}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\subsection{Climatic setting}
% ----------------------------------------------------------------------

@@

% ======================================================================
\section{Software tools and contributions}
% ======================================================================

@@

% ----------------------------------------------------------------------
\subsection{A Python PDD model}
% ----------------------------------------------------------------------
\todo{Figure. PyPDD demo example}

% ----------------------------------------------------------------------
\subsection{Pre-processing scripts for GRASS GIS}
% ----------------------------------------------------------------------

% ----------------------------------------------------------------------
\subsection{Plotting tools for PISM}
% ----------------------------------------------------------------------
\todo{Figure. Plotting tools example}

@@

>>>>

% My commands
\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
\newcommand{\todo}[1]{\emph{[\textbf{Todo:} #1]}}
\newcommand{\mref}[0]{\textbf{[ref.]}}

@@

% Document properties
\title{Numerical modelling of the Cordilleran ice sheet}
%\runningtitle{Numerical modelling of the Cordilleran ice sheet}
\author{Julien Seguinot}

@@

\newpage
%\maketitle
\tableofcontents
\newpage

@@

In our daily life, we most commonly experience ice as small, solid cubes
that crackle on contact with refreshing liquids. Yet, on geological
time scales, and under high pressure, ice behaves quite differently.
In the cold and mountainous regions of the globe, ice accumulates through
cumulative snowfall until it reaches a critical thickness. Beyond this
threshold, usually in excess of 50\,m, but dependent on temperature, ice
deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
Glaciers, these bodies of moving ice, move by a combination of
meltwater-induced sliding at the base \citep[\S~532]{Saussure.1786}\renote{
    I mean volume 1, which as a couple of beautiful paragraphs (would be worth
    translating if I had time) about how the "heat from the Earth's interior"
    melts basal ice to lubricate glacier flow. This is precisely those
    paragraphs that \citet{Forbes.1846} refers to sixty years later when
    advancing his ``viscous theory of glacier notion''. According to the cover
    page, volume 1 was published in ``MDCCLXXXVI'', which I just translated to
    1786 (so 1796 was wrong).},
and viscous deformation within the
ice body \citep{Forbes.1846b}.

As glaciers flow and slide across their bed, they transport\renote{
    I would like to keep this statement general, surface transport is also
    important. Erosion is not only by basal rock debris but also, indirectly,
    by glacial meltwater.}
rock debris and erode the landscape, thereby leaving geomorphological traces of
their former presence (\cref{fig:glaciation-indicators}). In several parts of
the world, montane people\renote{
    I insist on keeping this: people of the Alps, for istance, had understood
    that glacial erratics originated from glaciers before it has ever been
    documented by explorers and scientists. It is stated in the old literature.
    Check the new reference, reprinted in \citet{Mathews.1898} (not quite sure
    how to cite this material).}
and early explorers learned since long to read these traces, and to
understand that glaciers had once been more extensive than today
\citep[e.g.,][p.~21]{Windham.Martel.1744}. Wtih the emergence of the glacial
theory came the idea that, under colder temperatures
than today, expansive \emph{ice sheets} had once covered much of Europe and
North America \citep{Agassiz.1840}\renote{
    I am a bit reluctant to write ``\citep{Agassiz.1840} eventually posed the
    glacial theory''. Although he published these ideas, do they really
    originate from him?}.

However, the glacial theory did not gain general acceptance until the
discovery and exploration
of the two present-day ice sheets on Earth: the Greenland and
Antarctic ice sheets, provided a modern analogue for the proposed European and
North American ice sheets. The first traverse of Greenland was performed by
Nansen in 1888, while Amundsen reached the south pole in 1911. Meanwhile, in
the European Alps, it was then discovered from observations of the glacial
landscape, that there had not been only one, but at least four, glacial periods
\citep{Penck.Bruckner.1909}.

More recently, palaeoclimate records extracted from deep sea sediments and ice
cores have provided a much more detailed picture of the Earth environmental
history \citep[e.g.,][]{Emiliani.1955, Shackleton.Opdyke.1973,
Dansgaard.etal.1993}, indicating neither one, nor four, but rather tens of
glacial cycles. During the last 800\,ka years, glacial and interglacial cycles
have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
Augustin.etal.2004}. In this thesis, I focus on the most recent of these
periods: \emph{the last glacial cycle} (\cref{fig:paleo-timeseries})

% ice volume 26.07 + 2.93 = 29.00
% meters sea-level 58.3 + 7.36 = 65.6
Together, the Greenland and Antarctic ice sheets currently host almost all
glacial ice on the planet, corresponding to 65.6\,m of potential sea level
rise, were these ice sheets to melt away entirely \citep{Bamber.etal.2013,
Fretwell.etal.2013}. At the apogee of the last glacial period, global sea level
was 120 to 135\,m lower than today
\citep{Clark.Mix.2002}, implying an additional, double amount of ice stored in
ice sheets and glaciers on land. Systematic mapping and compilation of glacial
landforms from all continents \citep[e.g.,][]{Prest.etal.1968,
Boulton.Clark.1990, Kleman.etal.1997, Hattestrand.1998} permitted
reconstructions of these former ice covers \citep{Ehlers.Gibbard.2007},
indicating that the excess ice volume
was repartitioned between more extensive Greenland and Antarctic ice sheets,
the Fennoscandian and Barents-Kara ice sheets in the Eurasian Arctic, and the
Innuitian, Laurentide, and \emph{Cordilleran} ice
sheets over North America (\cref{fig:paleo-glaciation}).

Nevertheless, the landform record is sparse in time and, most often, spatially
incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
lanscape, and much of this evidence has been overridden by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994,
Kleman.etal.2006, Kleman.etal.2010}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical ice sheet models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
help. Although these models alone contain too many uncertainties to reproduce
accurate ice sheet reconstructions during former glacial cycles
\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
data where they exist \citep{Napieralski.etal.2007}, while allowing
physics-based, quantitative reconstructions of ice sheet sectors where they do
not
\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
In addition, numerical ice sheet models provide quantitative insights on the
four-dimensional distribution of stress and deformation in palaeo-ice sheets
and allow for reconstructions their past thickness and volume evolution, which
are difficult to infer from geological evidence
\renote{
    Is this the place where to cite \citep{Denton.Hughes.1981}? \todo{I need to
    check the contents of this book (library).}
}.

In this thesis, I focus on numerical modelling of the last Cordilleran ice
sheet, which once covered much of the mountainous terrain of western North
America \citep{Dawson.1888}, a region where topographic complexity has
presented a particular challenge to reconstructing glacial history based on the
geomorphological record \citep[e.g.,][]{Kleman.etal.2010} or numerical ice
sheet modelling \citep{Robert.1991}.

@@

% ======================================================================
\section{Physical setting}
% ======================================================================

During the last glacial cycle, much of North America has been covered by ice
(\cref{fig:map-northamerica}).
During the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
(\cref{fig:map-northamerica}). The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, \emph{the Cordilleran ice sheet} capped the mountain
ranges of the Western Cordillera, which forms our region of focus in the
present thesis.

@@

% ----------------------------------------------------------------------
\subsection{Topography}
% ----------------------------------------------------------------------

@@

% ----------------------------------------------------------------------
\subsection{Climate}
% ----------------------------------------------------------------------

@@

% ======================================================================
\section{Software implementation}
% ======================================================================

@@

@@

% ----------------------------------------------------------------------
\subsection{Peripheral tools}
% ----------------------------------------------------------------------
A Python PDD model\\
Pre-processing scripts for GRASS GIS\\
Plotting tools for PISM\\
